---
title: "Week 3 Assignment"
author: "Aryeh S"
date: "3/6/2023"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
# NOTE: I used the patchwork package in this analysis. Please install it before 
# knitting the data
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(patchwork)
```

# NYC Shooting Data

The following data set contains information regarding shooting data in NY 
during the years of 2006-2021. We will look at some demographic information
regarding the victims and perpetrators.

## Data Import and preparation for analysis

```{r Import and Clean Data}
#load data 
shooting_data=read_csv("https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD")

# get a quick look
glimpse(shooting_data)

# remove unwanted columns
(shooting_data = shooting_data %>% 
  select(-c(INCIDENT_KEY,PRECINCT,JURISDICTION_CODE,
            X_COORD_CD,Y_COORD_CD,Latitude,Longitude,Lon_Lat,LOCATION_DESC)))

# see unique values for variables of interest and set values for NA accordingly
unique(shooting_data$PERP_AGE_GROUP)
shooting_data$PERP_AGE_GROUP = replace_na(shooting_data$PERP_AGE_GROUP,"UNKNOWN")

unique(shooting_data$PERP_SEX)
shooting_data$PERP_SEX = replace_na(shooting_data$PERP_SEX,"U")

unique(shooting_data$PERP_RACE)
shooting_data$PERP_RACE = replace_na(shooting_data$PERP_RACE,"UNKNOWN")

# change data types to appropriate type
shooting_data = shooting_data %>%
  mutate(OCCUR_DATE = parse_date(OCCUR_DATE, format = "%m/%d/%Y"),
         OCCUR_MONTH =format(as.Date(OCCUR_DATE, format="%Y-%m-%d"),"%m"),
         PERP_AGE_GROUP = factor(PERP_AGE_GROUP),
         PERP_SEX = factor(PERP_SEX),
         PERP_RACE = factor(PERP_RACE),
         VIC_AGE_GROUP =factor(VIC_AGE_GROUP),
         VIC_SEX = factor(VIC_SEX),
         VIC_RACE =factor(VIC_RACE),
         BORO=factor(BORO),
         Time_of_Day = as.integer(str_sub(shooting_data$OCCUR_TIME,start=1,end=2)),
        DAY_OF_WEEK =wday(OCCUR_DATE),
        YEAR=year(OCCUR_DATE)
         )%>%
# filter outliers  
         filter(PERP_AGE_GROUP!=940 & PERP_AGE_GROUP!=224 & PERP_AGE_GROUP!=1020) %>%
         filter(!grepl('AMERICAN INDIAN/ALASKAN NATIVE',PERP_RACE))%>%
         filter(!grepl('ASIAN / PACIFIC ISLANDER',PERP_RACE))%>%
         filter(!grepl('AMERICAN INDIAN/ALASKAN NATIVE',VIC_RACE))%>%
         filter(!grepl('ASIAN / PACIFIC ISLANDER',VIC_RACE))


# When we look at the data now, we see that it is ready for analysis!
summary(shooting_data)
```
## Data Visualization
```{r Analysis}
p1 = ggplot(shooting_data, aes(x=PERP_SEX))+ geom_bar(fill="blue")+
  labs(x="Sex of Perpetrator",y="Number of Occurences")
p2 = ggplot(shooting_data, aes(x=PERP_RACE))+ geom_bar(fill="red")+
  labs(x="Race of Perpetrator",y="Number of Occurences")
p3 = ggplot(shooting_data, aes(x=PERP_AGE_GROUP))+ geom_bar(fill= "orange")+
  labs(x="Age Group of Perpetrator",y="Number of Occurences")
p4 = ggplot(shooting_data, aes(x=VIC_SEX))+ geom_bar(fill="blue")+
  labs(x="Sex of Victim",y="Number of Occurences")
p5 = ggplot(shooting_data, aes(x=VIC_RACE))+ geom_bar(fill="red")+
  labs(x="Race of Victim",y="Number of Occurences")
p6 = ggplot(shooting_data, aes(x=VIC_AGE_GROUP))+ geom_bar(fill="orange")+
  labs(x="Age Group of Victim",y="Number of Occurences")

# Plotting Sex of Perpetrator and Victim
p1/p4
# Plotting Race of Perpetrator and Victim
p2/p5
# Plotting Race
p3/p6

```
In Plot #1 We can see that the vast majority of perpetrators and victims are male.
It is interesting that there many cases were the sex of the perpetrator is given
as unknown, but there are very few unknowns in the set of victims.
It is unclear why this is the case-perphaps the perpetrator
escaped the police and could not be identified-but it seems plausible to assume
that the significant majority of the unkown perpetrators were male

In plot #2, we again find many cases where the race of the perpetrator is unknown.
However, for the data we do have, we find that black population is highly 
over represented in both the perpetrator and victim groups.

In plot #3, we again find many cases where the age of the perpetrator is unknown,
but for the data we have, we can see that the bulk of victims and perpetrators were 
between the ages of 18-44.

```{r Data Analysis }
#percentage of shootings that resulted in deaths
(percent_murders = mean(shooting_data$STATISTICAL_MURDER_FLAG))

#showing crimes that resulted in homicide vs only injured victim
ggplot(shooting_data,aes(x=STATISTICAL_MURDER_FLAG))+
  geom_bar(fill="purple")
#showing where the crimes took place
ggplot(shooting_data,
       aes(x=reorder(BORO,BORO,function(x)-length(x)))) +
  geom_bar(fill="green") +
  xlab("BORO")

#showing when the crimes took place
p1=ggplot(shooting_data,aes(x=DAY_OF_WEEK))+
  geom_bar(fill="blue")
p2=ggplot(shooting_data,aes(x=OCCUR_MONTH))+
  geom_bar(fill="red")
p3=ggplot(shooting_data,aes(x=Time_of_Day))+
  geom_bar(fill="orange")
p4=ggplot(shooting_data,aes(x=YEAR))+
  geom_bar(fill="purple")


p1
p2
p3
p4
```
We can see that:
1) Brooklyn had the most crimes
2) Crimes occured most frequently on the weekend and in the summer months
3) the proportion of shootings that results in deaths is ~19%
4) there was a general decrease in shootings from 2005-2020, followed by
a sharp uptick

## Bias Discussion

It is possible that the police were more or less active in certain areas and
with certain demographics, this would result in a disproportionate reporting
for certain categories of data. The population data for New York, which is not
contained in this dataset, may be inherently non-uniform. This would result
in different expected values for different demographics. If so, the skewed results
we saw in this dataset would not be unexpected. This would require a further detailed
analysis

```{r session info}
sessionInfo()
```


